---
title: "Collettorichum_phylogeny"
author: "Emily Giroux"
date: "2023-01-30"
output: html_document
fontsize: 11pt
geometry: margin=1in
urlcolor: blue
header-includes: \usepackage{xcolor}
---

## __Colletotrichum Phylogenomics Analysis__  
  
#### __*Project Name Setting*__
```{r}
analysis <- "Colletotrichum_phylogeny"
```
  
#### __*Set up the project's global options*__  
```{r, global_options, eval=TRUE, echo=FALSE, cache=TRUE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy = TRUE, fig.align='center',
               cache=FALSE, collapse=TRUE, echo=FALSE, eval=FALSE, include=FALSE,
               message=FALSE, quietly=TRUE, results='hide', warn.conflicts=FALSE, 
               warning=FALSE)
```
  
#### __*Project directory set-up*__  
Use the rprojroot package to set the directory structures and create directories    
that will be used throughout the script. The root object contains a function that    
will help us locate our package R files regardless of our current working directory.
```{r sourcing_my_functions, echo=FALSE, eval=TRUE, include=FALSE, cache=TRUE}
library("rprojroot")
root <- rprojroot::is_rstudio_project

# Capture, or create if not yet there, directory for misc. R script files:
scriptsPath <- root$make_fix_file(".")("R")
scripts  <- dir(root$find_file("R", path = root$find_file()))
scriptsList <- paste(scriptsPath, scripts, sep = "//")
lapply(scriptsList, source)

# Capture, or create if not yet there, directory for misc. Python script files:
scriptsPyPath <- root$make_fix_file(".")("pythonScripts")
dir.create(scriptsPyPath, showWarnings = TRUE, recursive = TRUE)
scriptsPy <- dir(root$find_file(scriptsPyPath, path = root$find_file()))
scriptsPyList <- paste(scriptsPyPath, scriptsPy, sep = "//")

# Create a directory for results and outputs - I call this the working directory:
sharedPath <- root$make_fix_file(".")("PIRL_working_directory")
dir.create(sharedPath, showWarnings = TRUE, recursive = TRUE)
sharedPathDirs <- dir(root$find_file(sharedPath, path = root$find_file()))
sharedPathAn <- paste(sharedPath, analysis, sep = "/")
dir.create(sharedPathAn, showWarnings = TRUE, recursive = TRUE)

# Create directory for genome databases:
databasePath <- root$make_fix_file(".")(paste("Databases", analysis, sep = "/"))
dir.create(databasePath, showWarnings = TRUE, recursive = TRUE)
databasesPathDir <- dir(root$find_file(databasePath, path = root$find_file()))

# Create separate database sub-directories for RefSeq, GenBank and Ensembl:
refSeqDB <- paste(databasePath, "refSeq", sep = "/")
gbDB <- paste(databasePath, "genbank", sep = "/")
ensemblDB <- paste(databasePath, "ensembl", sep = "/")
subfolder_names <- c("/refSeq", "/genbank", "/ensembl")
for (j in 1:length(subfolder_names)){
  folder <- dir.create(paste0(databasePath, subfolder_names[j]), showWarnings = TRUE, recursive = FALSE)
}

# Create separate database directories for genomes of species outside main 
# targets set, i.e., outgroups or other species within genus to improve phylogeny
# tree resolution: 
generaGenomes <- paste(databasePath, "all_genomes_under_target_genera", sep = "/")
refSeqDBGenus <- paste(generaGenomes, "refSeq", sep = "/")
gbDBGenus <- paste(generaGenomes, "genbank", sep = "/")
ensemblDBGenus <- paste(generaGenomes, "ensembl", sep = "/")
for (j in 1:length(subfolder_names)){
  folder <- dir.create(paste0(generaGenomes, subfolder_names[j]), showWarnings = TRUE, recursive = TRUE)
}

# Create a directory to hold all the saved environment files:
imagePath <- root$make_fix_file(".")(paste("r_environments", analysis, sep = "/"))
dir.create(imagePath, showWarnings = TRUE, recursive = TRUE)
images <- dir(root$find_file(imagePath, path = root$find_file())) # if there is already a data file in this directory if not running this for the first time
imagesList <- paste(imagePath, images, sep = "//")
load(paste(imagesList, sep = "/")) # Need to be specific if multiple images are present.
```
  
#### __*Project environment quick-load*__  
```{r}
library("rprojroot")
root <- rprojroot::is_rstudio_project
imagePath <- root$make_fix_file(".")(paste("r_environments", analysis, sep = "/"))
imageA <- "colleto_phylo.RData" # should provide this variable in initial chunk where image directory is specified - but concern is that there may be the need to divide and create different files within the same anaylsis....
save.image(paste(imagePath, imageA, sep = "/")) # you need to bget more specific if you have more than one image saved in this directory
```
  
### Project Steps:  
#### 1. Core project genomic information. Collect file location information for our primary 9 genomes.  
#### 2. Extend genome data to include other Colletotrichum species, using curated reference genomes from RefSeq.  
#### 4. If more genomic data is required, consider non-RefSeq data.  
#### 3. Create phylogeny from all the available genomes gathered.     
  
#### __Step 1__  
  
__*Collect paths to our personally-sequenced Colletotrichum genomes:*__  
Import csv that has our reference genomes directory paths and shortnames.
```{r}
sequencedColletoGenomes <- read.csv(file = root$find_file("metadata_colletoMar2022.csv", path = root$find_file()))
```
  
#### __Step 2__  
#####__Search for other quality Colletotrichum reference genomes available__  
__*a)*__ Use nearest_genomes.py script to gather species taxIDs. First need to get TaxIDs
of Colletotrichum species that have sequenced genomes. To get these, use the genus
TaxID as a starting point and then use the nearest_genomes python script to retrieve species TaxIDs.
  
__*b)*__ Use the species taxID information to gather additional information related to
genomic data, i.e., assembly and accession info for genome data retrieval.  
  
__*c)*__ Use the accessions or assembly IDs to retrieve genomic data using biomartr to retrieve data from on NCBI, RefSeq and Ensembl.  
  
  
__*Capture the Colletotrichum genus TaxID:*__
```{r}
target <- "5455" # genus tax-id for Colletotrichum
```
  
__*Prepare for function "nearest_genomes.py":*__  
**Unlike the APS project, we don't have a list from a table to go through, but I made it like this in case an outgroup genus is to be added later.**
```{r}
library("data.table")
generaList <- c("Colletotrichum") 
nearestGensTbl <- as.data.table(generaList)
```
  
__*Customize the nearest_genomes.py script to search using our target genus*__  
```{r}
library("reticulate")
nearestGenPyPath <- paste(scriptsPyPath, scriptsPy, sep = "//")
# At line 84 in nearest_genomes.py, replace "Aleurina" with the genus you want to seach, then save.
source_python(nearestGenPyPath)

for(i in 1:nrow(nearestGensTbl)){
  accns_r <- nearest_genomes(nearestGensTbl$generaList[i])
  nearestGensTbl$accns[i] <- paste(accns_r, collapse = ", ")
}

nearestGensTbl <- nearestGensTbl[, .(accns = unlist(tstrsplit(accns, ",", type.convert = TRUE))), by = generaList]
nearestGensTbl$accns <- sub(" ", "", nearestGensTbl$accns)
```
  
__*Save the environemnt image:*__
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```

Getting reference genomes available under the Colletotrichum genus, here we use Ensembl Genome Retrieval: (Run as background job)
```{r}
library(biomartr)
for(i in 1:nrow(nearestGensTbl)){
  biomartr::getGenomeSet(db = "ensembl", nearestGensTbl$accns[i], reference = FALSE, release = NULL,
                         clean_retrieval = TRUE, gunzip = FALSE, update = FALSE, path = ensemblDBGenus,
                         assembly_type = "toplevel")
    biomartr::getAssemblyStats(db = "ensembl", nearestGensTbl$accns[i], reference = FALSE, 
                             type = "download", 
                             path = ensemblDBGenus)
  warnings()
}

for(i in 1:nrow(nearestGensTbl)){
  biomartr::getGenomeSet(db = "genbank", nearestGensTbl$accns[i], reference = FALSE, release = NULL,
                         clean_retrieval = TRUE, gunzip = FALSE, update = FALSE, path = gbDBGenus,
                         assembly_type = "toplevel")
    biomartr::getAssemblyStats(db = "genbank", nearestGensTbl$accns[i], reference = FALSE, 
                             type = "download", 
                             path = gbDBGenus)
  warnings()
}

### Over here!!!! 21Feb2023 - above failed to retrieve genomes in list under genus - need to see what is going on with the fucking code.

for(i in 1:nrow(targetMeta)){
  res <- biomartr::is.genome.available(db = "genbank", organism = targetMeta$SpeciesName[i])
  targetMeta$genDB[i] <- res
}

gbGens <- targetMeta[targetMeta$genDB == "TRUE",]
for(i in 1:nrow(gbGens)){
  res <- biomartr::is.genome.available(db = "genbank", organism = gbGens$SpeciesName[i], details = TRUE)
  gbGens$assmblyAccn[i] <- res
}

library(biomartr)
for(i in 1:nrow(gbGens)){
  biomartr::getGenomeSet(db = "genbank", gbGens$SpeciesName[i], reference = FALSE, release = NULL,
                         clean_retrieval = TRUE, gunzip = FALSE, update = FALSE, path = gbDB,
                         assembly_type = "toplevel")
    biomartr::getAssemblyStats(db = "genbank", gbGens$SpeciesName[i], reference = FALSE, 
                             type = "download", 
                             path = gbDB)
  warnings()
}
```

Save env image:
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```


Obtain genomes from all species within a genus available from database:
```{r}
getKingdoms(db = "refseq")
getGroups(db = "refseq", kingdom = "fungi")
```













Need to obtain reference genomes for the target pathogens in the target Meta table:
```{r}
library("biomartr")
# set default timeout setting from 60s to at least 3000000 s before running any retrieval functions:
options(timeout = 30000000)
# List available database:
biomartr::listNCBIDatabases(db = "all")
# also try BarcodingR: https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12682
```

RefSeq Genome Check:
```{r}
library("biomartr")
res <- biomartr::is.genome.available(db = "refseq", organism = "Colletotrichum", details = TRUE)
# notice that the last 3 rows/entries are not from Colletotrichum genus, rather they are viral, so remove these from the table:
res <- res[1:14,]
```
Save env image:
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```


```{r}
# targetMeta$genRefSeq[i] <- res
refSeqGens <- res
refSeqGens$genRefSeq <- "TRUE"
# refSeqGens <- targetMeta[targetMeta$genRefSeq == "TRUE",]
for(i in 1:nrow(refSeqGens)){
  res <- biomartr::is.genome.available(db = "refseq", organism = refSeqGens$organism_name[i], details = TRUE)
  refSeqGens$assmblyAccn[i] <- res
}
```
RefSeq Genome Download of most recent version. Set up as a background job (Don't copy results - just double-check download folder has all genomes downloaded):
```{r}
options(timeout = 300000000)
library(biomartr)
for(i in 1:nrow(refSeqGens)){
  biomartr::getGenomeSet(db = "refseq", refSeqGens$organism_name[i], reference = FALSE, release = NULL, 
                         clean_retrieval = TRUE, gunzip = FALSE, update = FALSE, path = refSeqDB, 
                         assembly_type = "toplevel")
  biomartr::getAssemblyStats(db = "refseq", refSeqGens$organism_name[i], reference = FALSE, 
                             type = "download", 
                             #path = paste(refSeqDB, "/genomeassembly_stats", sep = ""))
                             path = refSeqDB)
  warnings()
}
```

Save env image:
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```

GenBank Genome Check:
```{r}
library("biomartr")
res <- biomartr::is.genome.available(db = "genbank", organism = "5455", details = TRUE)
# notice that the last 3 rows/entries are not from Colletotrichum genus, rather they are viral, so remove these from the table:
#res <- res[1:14,]
```
Save env image:
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```


```{r}
# targetMeta$genRefSeq[i] <- res
refSeqGens <- res
refSeqGens$genRefSeq <- "TRUE"
# refSeqGens <- targetMeta[targetMeta$genRefSeq == "TRUE",]
for(i in 1:nrow(refSeqGens)){
  res <- biomartr::is.genome.available(db = "refseq", organism = refSeqGens$organism_name[i], details = TRUE)
  refSeqGens$assmblyAccn[i] <- res
}
```
RefSeq Genome Download of most recent version. Set up as a background job (Don't copy results - just double-check download folder has all genomes downloaded):
```{r}
options(timeout = 300000000)
library(biomartr)
for(i in 1:nrow(refSeqGens)){
  biomartr::getGenomeSet(db = "refseq", refSeqGens$organism_name[i], reference = FALSE, release = NULL, 
                         clean_retrieval = TRUE, gunzip = FALSE, update = FALSE, path = refSeqDB, 
                         assembly_type = "toplevel")
  biomartr::getAssemblyStats(db = "refseq", refSeqGens$organism_name[i], reference = FALSE, 
                             type = "download", 
                             #path = paste(refSeqDB, "/genomeassembly_stats", sep = ""))
                             path = refSeqDB)
  warnings()
}
```

Save env image:
```{r}
save.image(paste(imagePath, imageA, sep = "/"))
```

Look at CAZyme distributions
```{r}
install.packages("readxl")
library("readxl")
cazyTblPath <- "/isilon/cfia-ottawa-fallowfield/users/girouxeml/PIRL_working_directory/Colletotrichum_phylogeny/CAZyme_prediction_result.xlsx"
cazyXLS <- read_excel(cazyTblPath)
```
```{r}

```


##################################################### OLD CODE NOT USED ANYMORE #######################


